\section{Context}

\college College is a small (1700 students), rural, highly selective
libreral arts college.  Students have generally shown academic
success; For example, the 75th percentile scores are 33 (ACT) and
1540 (new SAT composite); the 25th percentile scores are 30 and
1370.  Like many institutions, we have seen our program expand
rapidly.  During the time of this project, we have gone from
graduating between twelve and fifteen CS majors per year to graduating
nearly sixty CS majors each year.

Like many liberal arts colleges, \college considers the major a
small part of the overall undergraduate experience.  \college expects
a relatively lean curricular for majors; majors requirements may
not consist of more than eight courses plus a few non-departmental
requirements.\footnote{Students can certainly do more than the
minimal requirements.  But the requirements cannot exceed eight
courses in the department.} Our current departmental curricular
requirements include of a three-course, multi-language, introductory
sequence; systems (either architecture or operating systems);
upper-level algorithm design and analysis; theory of computation;
software design; and one departmental elective.  CS majors must
also take three semesters of mathematics and statatistics, including
Calculus, discrete mathematics, and a third course of the student's
choice.  

In spite of these limitations, we provide our students with a strong
and comprehensive undergraduate aducation, enough so that \college's
CS curriculum serves as one of the curricular exemplars in the
recent ACM/IEEE Curricular Guidelines \cite{curriclum-2013}.

Nonetheless, the Software Design curriculum has been particularly
problematic.  As one colleague put it, ``Software Design often feels
like a bunch of disconnected topics placed together into a single
course, with a large, ambiguous project tied on.''  Throughout the
years, the course has been offered in multiple forms, with different
languages, different emphases, and different textbooks.  In most
cases, the experience was that students did not take what we hoped
out of the course and that the course received some of the lower
end-of-course ratings in the department, no matter who taught it.
Even when a model seemed successful for one semester, subsequent
offerings in the same model showed less success.

